WineQualityReds by Farnaz_Motamediyan
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(MASS)
library(scales)

```


```{r echo=FALSE, Load_the_Data}
#setwd("~/Developer/data-analysis/EDA with R")
RedWineData= read.csv("wineQualityReds.csv")
names(RedWineData)
str(RedWineData)
nrow(RedWineData)
ncol(RedWineData)
summary(RedWineData)
summary(RedWineData$quality)
```

> 
RedWineData dataset has 13 variables of 1599 entries:
"fixed.acidity","volatile.acidity","citric.acid","residual.sugar","chlorides",
"free.sulfur.dioxide","total.sulfur.dioxide","density","pH","sulphates","alcohol",
"quality"
Quality is an ordinal variable with possible ranking from 1 (worst) to 10 (best).

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots1}

univ_cont <- function(feature) {
    grid.arrange(
    ggplot(data=RedWineData, aes_string(x = feature)) + geom_histogram(bins = 30),
    ggplot(aes_string(x=1, y=feature), 
        data = RedWineData) +
  geom_boxplot( )  , nrow =1)
}

fa <- univ_cont("alcohol")
#histogram and boxplot of the feature alcohol!Univariate_Plots.
```

histogram and boxplot of the level of alcohol in our red wine dataset!
Alcohol distribution is slightly skewed. the mean and median values of 
alcohol distribution are almost same.
The minimum of alcohol in all wines on our dataset is 8%.

```{r echo=FALSE, Univariate_Plots2}
fa <- univ_cont("quality")
```

histogram and boxplot of the quality in our red wine dataset!

```{r echo=FALSE, Univariate_Plots3}
fa <- univ_cont("sulphates")
```

histogram and boxplot of the sulphates in our red wine dataset!
Density and sulphates ditributions, like others, has long tails.

```{r echo=FALSE, Univariate_Plots4}

fa <- univ_cont("pH")
```

histogram and boxplot of the pH in our red wine dataset!

```{r echo=FALSE, Univariate_Plots5}

fa <- univ_cont("density")
```

histogram and boxplot of the density in our red wine dataset!
Density and sulphates ditributions, like others, has long tails.

```{r echo=FALSE, Univariate_Plots6}

fa <- univ_cont("total.sulfur.dioxide")
```

histogram and boxplot of the total.sulfur.dioxide in our red wine dataset!

```{r echo=FALSE, Univariate_Plots7}

fa <- univ_cont("free.sulfur.dioxide")
```

histogram and boxplot of the free.sulfur.dioxide in our red wine dataset!

```{r echo=FALSE, Univariate_Plots8}

fa <- univ_cont("free.sulfur.dioxide")
```

histogram and boxplot of the free.sulfur.dioxide in our red wine dataset!

```{r echo=FALSE, Univariate_Plots9}

fa <- univ_cont("chlorides")
```

histogram and boxplot of the chlorides in our red wine dataset!

```{r echo=FALSE, Univariate_Plots10}

fa <- univ_cont("residual.sugar")
```

histogram and boxplot of the residual.sugar in our red wine dataset!

Residual sugar is highly positively skewed. 
In addition, the plot contains two peaks, it is visioble commonly in
lots of plots, 
could be mainly due to wine type.

```{r echo=FALSE, Univariate_Plots11}

fa <- univ_cont("chlorides")
```

histogram and boxplot of the chlorides in our red wine dataset!

```{r echo=FALSE, Univariate_Plots12}

fa <- univ_cont("citric.acid")
```

histogram and boxplot of the citric.acid in our red wine dataset!
Citric acid distribution looks slightly bimodal and there are few outliers 
as well. one intresting thing that i noticed is unsual spikes 
around 0.0 g/dm^3 and 0.5 g/dm^3, this may indicate few concentrations 
are more common than others.

```{r echo=FALSE, Univariate_Plots13}

fa <- univ_cont("volatile.acidity")

```

histogram and boxplot of the volatile.acidity in our red wine dataset!
Both distributions for fixed acidity and volatile acidity have long positive 
tails, this makes their mean higher than their medians, and make median
better measure of central value.

```{r echo=FALSE, Univariate_Plots14}

fa <- univ_cont("fixed.acidity")

```


As it is illustrated above in the charts and histograms of variables, there are
some outliners that stand out in the chart.
Both distributions for fixed acidity and volatile acidity have long positive 
tails, this makes their mean higher than their medians, and make median
better measure of central value.
Citric acid distribution looks slightly bimodal and there are few outliers 
as well. one intresting thing that i noticed is unsual spikes 
around 0.0 g/dm^3 and 0.5 g/dm^3, this may indicate few concentrations 
are more common than others.
Residual sugar is highly positively skewed. 
In addition, the plot contains two peaks, it is visioble commonly in
lots of plots, 
could be mainly due to wine type.
Density and sulphates ditributions, like others, has long tails. 
Alcohol distribution is slightly skewed. the mean and median values of 
alcohol distribution are almost same.
The minimum of alcohol in all wines on our dataset is 8%.

We can even remove outliers if we find it appropriate, 
that will make the following analysis more robust.
http://www.public.iastate.edu/~maitra/stat501/lectures/Outliers.pdf
However, at this stage I decided to include outliners in order for the reader
to have a better underestanding of our current dataset.




# Univariate Analysis

I myself am a Red Wine lover.
In this dataset first variable that caught my interest was the quality of wine. 
It is usually ranked from 1 to 10, however, in this dataset is ranked from 1 to
8. 
Here in the bellow histogram, 
I categorized the data set by good(>=7) average(5 =< Avg <7) and poor(<5) wine 
based on its quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}

RedWineData$qualityrating <- ifelse(RedWineData$quality < 5, 'poor', ifelse(
  RedWineData$quality < 7, 'average', 'good'))
RedWineData$qualityrating <- ordered(RedWineData$qualityrating,
                     levels = c('poor', 'average', 'good'))
summary(RedWineData$qualityrating)
qplot(RedWineData$qualityrating)
```

#Distributions and Outliers

Looking at the first rounds of histograms that 
I created to look at the variables,
it appears that Qualitatively, 
residual sugar and chlorides have extreme outliers.
Citric has a large number of zero values, however,
i'm wondering whether this is a case of non-reporting of values.
It looks like density and pH are normally distributed, with only few outliers.
sulfur dioxides, fixed and volatile acidity, sulphates,
and alcohol seem to be long-tailed.

Arguments Log10: log is a character string
which contains "x" if the x axis is to be logarithmic. here I used log10

```{r}

ggplot(data = RedWineData,
       aes(x = citric.acid)) +
  geom_histogram() 
  
ggplot(data = RedWineData,
       aes(x = fixed.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(data = RedWineData,
       aes(x = volatile.acidity)) +
  geom_histogram() +
  scale_x_log10()

```

we can see that fixed.acidity and volatile.acidity 
appear to be normally-distributed.
 

### 
I created a new variable as qualityrating 
to rate the quality of wine as good average and poor.

### 

I addressed the distributions in the 'Distributions and outliers' section.
Here I will continue to visualize the outliers by Boxplots. 
But it is important to mention that at the end I did not perform any operations
on the data to tidy or adjust or change the form of the data here 
at Univariate Analysis section. However, 
I will do so with the Bivariable section.
In order to make visualization by boxplots, 
first I define a new function to create boxplot for each variable.



# Bivariate Plots Section

```{r echo=FALSE}

ggplot(data = RedWineData, aes(x = quality, y = alcohol,
                      fill = qualityrating)) +
  geom_boxplot() +
  ggtitle('percentage of Alcohol in Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')
```

In the above plot, the relation between alcohol and quality of red wine is illustrated.
Higher level of alcohol is positively correlated to better wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine_quality_groups= group_by(RedWineData,quality)
cor.test(RedWineData$quality ,RedWineData$fixed.acidity , method="pearson")
with(subset(RedWineData, quality > 6), cor.test(quality, fixed.acidity))
ggplot(aes(y=total.sulfur.dioxide, x=quality), data=RedWineData)+ 
  geom_jitter()
```


```{r echo=FALSE, Bivariate_Plots}

#here in the box plots we illustrate 
#the relationship between some features and quality of red wines.
#As it is shown, for example, quality increases as pH decreases or/and quality
#increases as alcohol increases.

plo1= qplot(x= qualityrating, y=fixed.acidity, 
            data= RedWineData, geom="boxplot" )
plo2= 
  qplot(x= qualityrating, y=volatile.acidity, 
        data= RedWineData, geom="boxplot")
plo3= 
  qplot(x= qualityrating, y=pH, data= RedWineData, geom="boxplot" )
plo4= 
  qplot(x= qualityrating, y=density, data= RedWineData, geom="boxplot" )
plo5= 
  qplot(x= qualityrating, y=alcohol, data= RedWineData, geom="boxplot" )

grid.arrange(plo1,plo2,plo3,plo4,plo5, ncol=4)


```

```{r echo=FALSE, Bivariate_Plots2}

biva_cont <- function(feature1,feature2) {
    grid.arrange(
    ggplot(aes_string(feature1, 
            feature2), 
        data = RedWineData ) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4))
}


fa <- biva_cont('factor(quality)', 
            'alcohol')

fa <- biva_cont("factor(quality)", 
            "pH")

fa <- biva_cont("factor(quality)", 
            "fixed.acidity")

fa <- biva_cont("factor(quality)", 
            "volatile.acidity")


fa <- biva_cont("factor(quality)", 
            "volatile.acidity")


```

# Bivariate Analysis

It showes at the above plots that a good wine has a lower pH , higher alcohol
and higher acidic (all three acidic kind examined here),
as an example.


###
some statistics on quality of wines. the question to be answered is, 
which are the lements that have most influence in the quality of wine?
```{r echo=FALSE, message=FALSE, warning=FALSE}

correlations_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}

correlations <- c(
  correlations_test(RedWineData$fixed.acidity, RedWineData$quality),
  correlations_test(RedWineData$volatile.acidity, RedWineData$quality),
  correlations_test(RedWineData$citric.acid, RedWineData$quality),
  correlations_test(log10(RedWineData$residual.sugar), RedWineData$quality),
  correlations_test(log10(RedWineData$chlorides), RedWineData$quality),
  correlations_test(RedWineData$free.sulfur.dioxide, RedWineData$quality),
  correlations_test(RedWineData$total.sulfur.dioxide, RedWineData$quality),
  correlations_test(RedWineData$density, RedWineData$quality),
  correlations_test(RedWineData$pH, RedWineData$quality),
  correlations_test(log10(RedWineData$sulphates), RedWineData$quality),
  correlations_test(RedWineData$alcohol, RedWineData$quality))
names(correlations) <- 
  c('fixed.acidity', 'volatile.acidity', 
    'citric.acid', 'log10.residual.sugar',
'log10.chlordies', 'free.sulfur.dioxide',
'total.sulfur.dioxide', 'density', 'pH',
'log10.sulphates', 'alcohol')
correlations
```


and here is the result of correlations:
fixed.acidity     volatile.acidity          citric.acid log10.residual.sugar 
          0.12405165          -0.39055778           0.22637251           0.02353331 
     log10.chlordies  free.sulfur.dioxide total.sulfur.dioxide              density 
         -0.17613996          -0.05065606          -0.18510029          -0.17491923 
                  pH      log10.sulphates              alcohol 
         -0.05773139           0.30864193           0.47616632 
         
As it appears, alcohol, sulphates, volatile acidity and citric acide have the highest correlations to quality. 

# Multivariate Plots Section

```{r  Multivariate_Plots1}

grid.arrange(
  ggplot(data = RedWineData, 
                    aes(x = alcohol, y = fixed.acidity,
                                   color = factor(qualityrating))) + 
               ylab('Fixed Acidity (g/dm^3)') +
               xlab('alcohol') +
               geom_boxplot(),
  
  ggplot(data = RedWineData, 
                    aes(x = alcohol, y = volatile.acidity,
                                   color = factor(qualityrating))) +
               ylab('Volatile Acidity (g/dm^3)') +
               xlab('alcohol') +
               geom_boxplot(), 
  
  ggplot(data = RedWineData,
                    aes(x = alcohol, y = citric.acid,
                                   color = factor(qualityrating))) +
               ylab('Citric Acid (g/dm^3)') +
               xlab('alcohol') +
               geom_boxplot(), 
  
  ggplot(data = RedWineData, 
                    aes(x = alcohol, y = pH,
                                   color = factor(qualityrating)) )+
               ylab('pH') +
               xlab('alcohol') +
               geom_boxplot())

```

In the above Multivariate plots, three factors been examined together.
I have choosen alcohol and combined it with one more component in order to examine
their combined impact on the quality of the wine.
For example, it is shown that higher acidity or lower pH 
is recognized in higher quality wines.
There are four colorful plots that illustrate correlation between 
the quality of wine and combination of alcohol and PH,Citric Acid,Fixed Acidity
and Volatile Acidity.


```{r echo=FALSE, Multivariate_Plots}

ggplot(data = RedWineData,aes(x = citric.acid, 
                        y = volatile.acidity,color = qualityrating)) +geom_point()

ggplot(data = RedWineData,
       aes(x = alcohol, y = log10(sulphates),
           color = qualityrating)) +
  geom_point()

ggplot(data = RedWineData,
       aes(x = pH, y = alcohol, color = qualityrating)) +
  geom_point()

 ggplot(aes(x = alcohol, 
           y = sulphates  , color = factor(qualityrating)), 
       data = RedWineData) +
      geom_point(alpha = 0.8, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
 

#relationship between volatile.acidity, alcohol and quality 
 ggplot(aes(x = alcohol, 
           y = volatile.acidity  , color = factor(qualityrating)), 
       data = RedWineData) +
      geom_point(alpha = 0.8, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
 
```

# Multivariate Analysis
I tried to identify and illustrate the 4 most correlated features to the quality of wines. As it is shown, higher citric acid and lower volatile acid could produce a better (in terms of quality) wine. lower pH , higher sulphates and higher alcohol(%) also shared contribution to a more high-quality wine.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}

  ggplot(data = RedWineData, 
                    aes(x = alcohol, y = fixed.acidity,
                                   color = factor(qualityrating))) + 
               ylab('Fixed Acidity (g/dm^3)') +
               xlab('alcohol') +
               geom_point(alpha = 0.8, size = 2)
  
  ggplot(data = RedWineData, 
                    aes(x = alcohol, y = volatile.acidity,
                                   color = factor(qualityrating))) +
               ylab('Volatile Acidity (g/dm^3)') +
               xlab('alcohol') +
               geom_point(alpha = 0.8, size = 2)
  
  ggplot(data = RedWineData,
                    aes(x = alcohol, y = citric.acid,
                                   color = factor(qualityrating))) +
               ylab('Citric Acid (g/dm^3)') +
               xlab('alcohol') +
               geom_point(alpha = 0.8, size = 2)

 ggplot(data = RedWineData, 
                    aes(x = alcohol, y = pH,
                                   color = factor(qualityrating)) )+
               ylab('pH') +
               xlab('alcohol') +
               geom_point(alpha = 0.8, size = 2)



```

### Description One
As we have identified it before, It is very interesting to observe
the relationship between higher acidity or lower pH with the
higher quality of wines.
There are four colorful plots that illustrate correlation between 
the quality of wine and alcohol and PH,Citric Acid,Fixed Acidity
and Volatile Acidity.
in the following plots we will break the plots down into
different correlations in order to underestand it better.
In the first following plot, we start with the alcohol and quality of wine.
and in the last plot we look and analyse an example of correlations which is alcohol and volatile acidity and their effect on the quality of wine.

### Plot Two

```{r echo=FALSE, Plot_Two}

ggplot(data = RedWineData, aes(x = quality, y = alcohol,
                      fill = qualityrating)) +
  geom_boxplot() +
  ggtitle('percentage of Alcohol in Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')
```

### Description Two

Here in these three boxplots we can see the effect of alcohol in the quality 
of wine. I personaly don't like the argument of ' higher alcohol 
results in higher quality of wine'. In fact I argue that the combination of
higher alcohol and other factors will produce higher qualities, 
and in  here this case is the acidic combination.
and the outliers here also showes that alcohol by itself alone
is not a strong indicator of high quality wine.

### Plot Three


```{r echo=FALSE, Plot_Three}


ggplot(data = subset(RedWineData ,qualityrating != 'average'), 
       aes(x = volatile.acidity,
            y = alcohol, color = factor(qualityrating))) +
  geom_point(alpha = 0.8, size = 2) + 
  geom_smooth(method = "lm", se = FALSE,size=1) + 
  ggtitle('Alcohol vs. Volatile Acidity and Wine Quality') + 
  xlab('Volatile Acidity (g / dm^3)') + ylab('Alcohol (% volume)')

```

### Description Three
I have done rating as good, average and poor. I named the new variable quality rating. in this scattered plot I neglected the average rate and only considered good and poor quality wines. I initially wanted to see the correlation of alcohol and volatile acidity on the quality of wine. As it is visible, higher volatile acidity brings down the quality of wine. Higher quality wines also tend to have higher alcohol. as a result, i can see that higher percentage of alcohol combined with lower volatile acidity produced a better wine in terms of quality.
Here I will try to visualise the correlation of each of my variables(in the first plot) with quality in order to support my argument.



# Reflection

This EDA or exploratory data analysis helped me gain insights 
about the redwinequality dataset. 
i was able to visualize relationships and correlations of different variables. 
I was able to identify the most related variables to the quality of red wine.
As it appears, alcohol, sulphates, 
volatile acidity and citric acide have the highest correlations
to the quality of red wines. 
I am interested to discover quality patterns of white wines to see
if they share some similarities. 

Extra information on Variable description

Description of attributes:

Fixed acidity: most acids involved with wine or fixed or nonvolatile
(do not evaporate readily)
Volatile acidity: the amount of acetic acid in wine, 
which at too high of levels can lead to an unpleasant, vinegar taste
Citric acid: found in small quantities, citric acid can add ???freshness??? 
and flavor to wines
Residual sugar: the amount of sugar remaining after fermentation stops, 
it???s rare to find wines with less than 1 gram/liter and wines with greater 
then 45 grams/liter are considered sweet
Chlorides: the amount of salt in the wine
Free sulfur dioxide: the free form of SO2 exists in equilibrium between
molecular SO2 (as a dissolved gas) and bisulfite ion; 
it prevents microbial growth and the oxidation of the wine
Total sulfur dioxide: amount of free and bound forms of S02; 
in low concentrations, SO2 is mostly undetectable in wine,
but at free SO2 concentrations over 50 ppm, SO2 becomes evident
in the nose and taste of wine
Density: the density of water is close to that of water
depending on the percent alcohol and sugar content
pH: describes how acidic or basic a wine is on a scale 
from 0 (very acidic) to 14 (very basic); most wines are 
between 3-4 on the pH scale
Sulphates: a wine additive which can contribute 
to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant
Alcohol: the percent alcohol content of the wine
Quality: output variable (based on sensory data, score between 0 and 10)

I also would like to expand a little bit on the struggles and successes
through the analysis.
working with R studio was indeed the easiest project throughout this Nanodegree.
I had a very smooth time completing this
project specially compare to OpenStreetMap. Having said that,
there were few logical arguments that I needed to read extensively
on the internet or receive help from mentors to understand. 
especially using factor element (color = factor(quality)) for the Multivariate
the analysis helped me a lot to better describe my plots and
better explain my arguments.

resources:
https://www.bbr.com/wine-knowledge/faq-quality
https://en.wikipedia.org/wiki/Red_wine
https://www.rstudio.com